---
title: "MG_Github"
output: html_document
date: "2026-05-11"
---

loading libraries
```{r}
library(stringr)
library(dplyr)
library(stats)
library(ggplot2)
library(ggpubr)
library(vegan)
library(cowplot)
library(tidyverse)
library(phyloseq)
library(microbiome)
library("funfuns")
library(pairwiseAdonis)
library(viridis)
library("microViz")
library(ggplot2)
library(cowplot)
library(phyloseq)
library(car)
library(ggpubr)
library(vegan) #vegan 2.6-4
library(dada2)
library(RColorBrewer)
library(dplyr)
library(tidyverse)
```

```{r}
getwd()
```

read in clean phyloseq object & seqtab object from dada2pipeline 

uploading 2023/2025 ps.cleaner data
```{r}
mg23.ps.cleaner <- readRDS("mgpt1.ps.cleaner .rds")
mg25.ps.cleaner <- readRDS("mgpt2.ps.cleaner.rds")
```

```{r}
sample_sums(mg23.ps.cleaner)[1:10]
sample_sums(mg25.ps.cleaner)[1:10]
```
applying a CLR (Centered Log-Ratio) transformation 
making a sequence table 
```{r}
mg23.ps.cleaner.clr <- microbiome::transform(mg23.ps.cleaner,'clr')
mg25.ps.cleaner.clr <- microbiome::transform(mg25.ps.cleaner,'clr')
mg23seqtab.clr <- data.frame(otu_table(mg23.ps.cleaner.clr))
mg25seqtab.clr <- data.frame(otu_table(mg25.ps.cleaner.clr))
```

~~~~~~~~~~~~~~~~~~~~~~~ Marsh Grass Pt 1 (High Freq Transect ~ 2023) Stats~~~~~~~~~~~~~~~~~

Beta Dispersion/Diversity!

Separating out metdata from mg23ps.cleaner 
```{r}
mg23.samdf <- data.frame(mg23.ps.cleaner@sam_data)
```

beta dispersion
*sig between months
*not sig between transects
```{r}
mg23.dist.all1 <- vegdist(mg23seqtab.clr, method = "euclidean")
mg23.bet.all1 <- betadisper(mg23.dist.all1,mg23.samdf$Month)
mg23.bet.all1
anova(mg23.bet.all1)


mg23.bet.all2 <- betadisper(mg23.dist.all1,mg23.samdf$Transect)
mg23.bet.all2
anova(mg23.bet.all2)
```
Beta diversity & Comp Comp Stats/Figures: 
*Month sig
*Transect not sig
```{r}
mg23.samdf$Month <- as.factor(mg23.samdf$Month)
mg23.samdf$Transect <- as.factor(mg23.samdf$Transect)
adonis2(mg23.dist.all1 ~ Month, data=mg23.samdf, permutations=9999) 
adonis2(mg23.dist.all1 ~ Transect, data=mg23.samdf, permutations=9999) 
```

Figure 1A: Plot to visualize month differences in 2023 samples 
```{r}
mg.all.ord1 <- plot_ordination(
  mg23.ps.cleaner.clr,
  ordinate(mg23.ps.cleaner.clr, "PCoA", "euclidean"),
  color = "Month"
) +
  stat_ellipse(linewidth = .5) +
  geom_point(size = 1, alpha = 0.8) +
  theme_classic(base_size = 22) +
  theme(
    axis.text.x = element_text(size = 2),
    axis.text.y = element_text(size = 2)
  ) +
  scale_color_manual(
    name = "Month",
    values = c("slateblue2",  "#7FC8A3","#7A9CC9")
  )

 
mg.all.ord1
```
ACOMB-BC2 diff abundance!
*sensitivity scores low, so only looked at structural zero data 
*results not reported in manuscript 
```{r}
mg23.diff.abund.tran <- ancombc2(
  data = mg23.ps.cleaner,
  tax_level = "Family",
  fix_formula = "Transect",
  p_adj_method = "holm",
  group = "Transect",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(mg23.diff.abund.tran$zero_ind)
#write.csv(mg23.diff.abund.tran$zero_ind, "mg23tran_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(mg23.diff.abund.tran$pseudo_sens_tab)
#write.csv(mg23.diff.abund.tran$pseudo_sens_tab, "mg23tran_sens.csv", row.names = TRUE)
```

*sensitivity scores low, so only looked at structural zero data 
*results from this are reported in table 3
```{r}
mg23.diff.abund.month <- ancombc2(
  data = mg23.ps.cleaner,
  tax_level = "Family",
  fix_formula = "Month",
  p_adj_method = "holm",
  group = "Month",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(mg23.diff.abund.month$zero_ind)
#write.csv(mg23.diff.abund.month$zero_ind, "mg23month_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(mg23.diff.abund.month$pseudo_sens_tab)
#write.csv(mg23.diff.abund.month$pseudo_sens_tab, "mg23month_sens.csv", row.names = TRUE)
```

Alpha Diversity Stats!
~read in alpha diversity dataframes 

how did we make these? 
alpha_diversity<- data.frame(estimate_richness(ps.cleaner, split=TRUE, measures=c("Shannon","InvSimpson","Observed")))
alpha_diversity$Evenness <- alpha_diversity$Shannon / log(alpha_diversity$Observed)
^did this for each experiment and saved dataframes to be uploaded below 
ps.cleaner phyloseq object should not have transformed data (should use raw or minimally processed data for this)
*if included below then the model passes assumptions + significant results 


~ alpha diversity dataframe ~ also has nutrient metadata (added in excel) 
```{r}
mg23.alpha <- read.csv("mg_diversity_df copy 2.csv")
```

splitting dataframe by month:
```{r}
newdf_list <- split(mg23.alpha, mg23.alpha$Month)

# Access each subset dataframe by its corresponding month value
june_df <- newdf_list$June
july_df <- newdf_list$July
december_df <- newdf_list$December
```

Linear Mixed Effect Models (for each alpha div metric ~ Shan, InvSimp,Evenness, Richness) separate for each month (Dec negated due to small sampling size))
*repeated for each alpha div metric, only included stat sig ones

June Shannon Diversity Model 
*TOM
```{r}
library(lmerTest)

juneall <- lmer(Shannon ~ Transect + TOC + TOM + TN + (1|Mat),
                data = june_df, REML = FALSE)

summary(juneall)


# Plot residuals against fitted values
plot(residuals(juneall) ~ fitted(juneall))

# Shapiro-Wilk normality test
shapiro.test(residuals(juneall))

# Q-Q plot of residuals
qqnorm(residuals(juneall))
qqline(residuals(juneall))
plot_model(juneall, type = 'diag')
#passes assumptions


```

July Shannon Diversity Model 
*Transect 
*TOC
*TN
```{r}
julyall <- lmer(Shannon ~ Transect + TOC + TOM + TN +  (1|Mat), data = july_df, REML = FALSE)
summary(julyall)

# Plot residuals against fitted values
plot(residuals(julyall) ~ fitted(julyall))

# Shapiro-Wilk normality test
shapiro.test(residuals(julyall))

# Q-Q plot of residuals
qqnorm(residuals(julyall))
qqline(residuals(julyall))
plot_model(julyall, type = 'diag')
#passes assumptions
```

June Evenness Model 
*Transect
```{r}
junealleven <- lmer(even ~ Transect + TOM  + TOC + TN + (1|Mat), data = june_df, REML = FALSE)
summary(junealleven)

# Plot residuals against fitted values
plot(residuals(junealleven) ~ fitted(junealleven))

# Shapiro-Wilk normality test
shapiro.test(residuals(junealleven))

# Q-Q plot of residuals
qqnorm(residuals(junealleven))
qqline(residuals(junealleven))
plot_model(junealleven, type = 'diag')
#passes assumptions
```

July Richness Model 
*Transect
*TOC
*TN 
```{r}
julyallrich <- lmer(Observed ~ Transect + TOM  + TOC + TN + (1|Mat), data = july_df, REML = FALSE)
summary(julyallrich)

# Plot residuals against fitted values
plot(residuals(julyallrich) ~ fitted(julyallrich))

# Shapiro-Wilk normality test
shapiro.test(residuals(julyallrich))

# Q-Q plot of residuals
qqnorm(residuals(julyallrich))
qqline(residuals(julyallrich))
plot_model(julyallrich, type = 'diag')
#passes assumptions
```

Nutrient Correlations!
```{r}
mgpt1mdffam <- read.csv ("mgpt1mdffam.csv")
#how this dataframe was made is specified below for bar graphs**
```

Looking at greater than 2% relative abundant taxa selected (unknown family negated)
```{r}
# Assuming mdfphy.rel is your data frame
top_fam <- aggregate(Abundance ~ Family, data =mgpt1mdffam , FUN = sum)

# Sort the result in descending order
top_fam <- top_fam[order(-top_fam$Abundance), ]

# Calculate the total abundance
total_abundance <- sum(top_fam$Abundance)

# Calculate the percentage for each phylum
top_fam$Percentage <- (top_fam$Abundance / total_abundance) * 100

# Display the top phyla with their percentages
print(top_fam)
```
Calculating correlation coefficients based on Pearson
```{r}
# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)

#mg23.alpha
#mgpt1mdffam

# Merge dataframes 
merged.mgpt1 <- merge(mg23.alpha, mgpt1mdffam, by = "Sample")

#Define top families (this is taxa w > 2% relative abundance) ---
specific_family <- c("A4b", "Cyclobacteriaceae", "Desulfocapsaceae",
                     "Flavobacteriaceae","Halieaceae","Rhodobacteraceae",
                     "Saprospiraceae","Woeseiaceae", "Pirellulaceae")

# Filter for top families & remove NAs 
merged_clean <- merged.mgpt1 %>%
  filter(Family %in% specific_family) %>%
  drop_na(Abundance, TOC, TOM, TN)

#  Pivot nutrients to long format 
corr_df_long <- merged_clean %>%
  pivot_longer(
    cols = c(TOC, TOM, TN),
    names_to = "Variable",
    values_to = "Nutrient"
  )

# Calculate correlations + p-values per family × nutrient
corr_family <- corr_df_long %>%
  group_by(Family, Variable) %>%
  summarize(
    Correlation_Coefficient = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$estimate,
      NA
    ),
    
    P_value = ifelse(
      length(unique(Nutrient)) > 1,
      cor.test(Abundance, Nutrient, method = "pearson")$p.value,
      NA
    ),
    
    .groups = "drop"
  )
```

Figure 1B
```{r}
heatmap_topfamily2 <- ggplot(
  corr_family,
  aes(
    x = Variable,
    y = Family,
    fill = Correlation_Coefficient
  )
) +
  geom_tile(color = "white") +
  
  scale_fill_gradient2(
    low = "#3385BB",        # blue = negative correlation
    mid = "#F2F2F2",        # light neutral = zero
    high = "#49AB3C",       # green = positive correlation
    midpoint = 0,
    name = "Pearson's Correlation"
  ) +
  
  scale_y_discrete(position = "right") +
  
  labs(
    x = "Nutrients",
    y = "Top Families"
  ) +
  
  theme_minimal() +
  
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    
    # Remove gridlines
    panel.grid = element_blank(),
    
    # Keep black box around heatmap
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    ),
    
    legend.position = "right"
  )

heatmap_topfamily2
```

~sig stars added later~
corr_fam uploaded to github for easy readability of sig correlations 


~~~~~~~~~~~~~~~~~~~~Marsh Grass Pt 2 (Low Freq Transect ~ 2025) Stats~~~~~~~~~~~~~~~~~~~

Beta dipsersion & beta diversity!

Separating out metadata from mg23ps.cleaner 
```{r}
mg25.samdf <- data.frame(mg25.ps.cleaner@sam_data)
```

beta dispersion
*not sig between transects
```{r}
mg25.dist.all1 <- vegdist(mg25seqtab.clr, method = "euclidean")
mg25.bet.all1 <- betadisper(mg25.dist.all1,mg25.samdf$Transect)
mg25.bet.all1
anova(mg25.bet.all1)
```

beta diversity
*Transect not sig
```{r}
mg25.samdf$Transect <- as.factor(mg25.samdf$Transect)
adonis2(mg25.dist.all1 ~ Transect, data=mg25.samdf, permutations=9999) 
```

ACOMB-BC2 diff abundance!
*sensitivity scores low, so only looked at structural zero data 
*results not reported in manuscript 
```{r}
mg25.diff.abund.tran <- ancombc2(
  data = mg25.ps.cleaner,
  tax_level = "Family",
  fix_formula = "Transect",
  p_adj_method = "holm",
  group = "Transect",    # factor you are testing
  struc_zero = TRUE,        # detects structural zeros (presence/absence)
  neg_lb = TRUE,
  em_control = list(
    tol = 1e-5,
    max_iter = 100,
    sensitivity = TRUE      # runs subsampling stability check
  ),
  alpha = 0.05,
  global = TRUE             # adds global tests in addition to pairwise
)
```

```{r}
#structural zero results (need to clean up in excel)
View(mg25.diff.abund.tran$zero_ind)
#write.csv(mg25.diff.abund.tran$zero_ind, "mg25tran_strzeros.csv", row.names = TRUE)
#sensitivity results (scores are low, this is why we negate global and pairwise results and focus on the robust structural zero results above)
View(mg25.diff.abund.tran$pseudo_sens_tab)
#write.csv(mg25.diff.abund.tran$pseudo_sens_tab, "mg25tran_sens.csv", row.names = TRUE)
```

Alpha Diversity!
Dataframe for alpha diversity 2025 samples was created the same way as for 2023 samples (specific info available above)

~ alpha diversity dataframe ~ also has nutrient metadata (added in excel) 
```{r}
mg25.alpha <- read.csv("alpha.div.mgpt2.csv")
```

*Note: tried using mat as a random effect as done above, but ~ A singular fit usually means the random effect variance is essentially zero, suggesting the model cannot distinguish variation among mats from residual variation. 
~In practical terms, Mat is probably not needed here
~Switched to one way anovas for these analyses 
```{r}
library(lme4)
july1 <- lmer(Shannon ~ Transect + (1|Mat), data = mg25.alpha, REML = FALSE)
summary(july1)
```

One--Way ANOVA
~ assumptions passed, but not significant 
~ same model tested for the other alpha diversity metrics (inv simpson, richness, evenness): no sig effects found for any of these 
```{r}
anov1 <- aov(Shannon ~ Transect, data = mg25.alpha)
summary(anov1)

#assumptions
mg25.alpha$Transect <- as.character(mg25.alpha$Transect)
# Normality of residuals
shapiro.test(residuals(anov1))
# Homogeneity of variances
library(car)
leveneTest(Shannon ~ Transect, data =mg25.alpha)

```
~~~~~~~~~~~~~~~~~~~~~~~~Bar Plots 2023 Month/Transect & 2025 Transect!~~~~~~~~~~~~~~~~~~~~~~
SET UP (for reproducibility)

Need to sum groups (transect + month for 2023 samples/ just transect for 2025 samples)
```{r}
library(microshades)
mg23.ps.cleaner.rel <- transform_sample_counts(mg23.ps.cleaner, function(x) x / sum(x))
fam.aggreg1 <- tax_glom(mg23.ps.cleaner.rel, "Family")
mgpt1mdffam<- prep_mdf(fam.aggreg1, subgroup_level = "Family")
```

2023 Transect example below: 
*repeated for month for 2023 samples
*completed for transect for 2025 samples
```{r}
library(dplyr)
mgpt1fam_transect <- mgpt1mdffam %>%
  group_by(Transect, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
mgpt1fam_transect <- mgpt1fam_transect %>%
  group_by(Transect) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

mgpt1fam_transect$Family_grouped <- ifelse(
  grepl("_", mgpt1fam_transect$Family),
  "Assigned at Higher Taxonomy",
  ifelse(mgpt1fam_transect$Percentage < 1, "Rare Families", mgpt1fam_transect$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
mgpt1fam_transect_grouped <- mgpt1fam_transect %>%
  group_by(Transect, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- mgpt1fam_transect_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

mgpt1fam_transect_grouped$Family_grouped <- factor(
  mgpt1fam_transect_grouped$Family_grouped,
  levels = family_order
)
```


new dataframe ~ just july 2023 samples ~ looking at the transect families in just this month! 

```{r}
#separating july for transect comparison image
mgpt1mdffamjuly <- mgpt1mdffam %>%
  filter(!Month %in% c("June ", "December"))
```

```{r}
library(dplyr)
julymgpt1fam_transect <- mgpt1mdffamjuly %>%
  group_by(Transect, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop")

#convert to relative abundance within each transect
julymgpt1fam_transect <- julymgpt1fam_transect %>%
  group_by(Transect) %>%
  mutate(Percentage = (Abundance / sum(Abundance)) * 100)

#grouping rules (same as above) applied
#families less than 1% relatively abundant grouped in "rare families" category 
#families only assigned at higher taxonomic rankings (phyla, class, order, ect) grouped into "Assigned at Higher Taxonomy" category
#note: some families were "unknown" meaning silva database was unable to find a match, so there should also be an "unknown" category unless all samples had identification to the family level

julymgpt1fam_transect$Family_grouped <- ifelse(
  grepl("_", julymgpt1fam_transect$Family),
  "Assigned at Higher Taxonomy",
  ifelse(julymgpt1fam_transect$Percentage < 1, "Rare Families", julymgpt1fam_transect$Family)
)

#re-sum after grouping (bc grouped categories are now collapsed)
julymgpt1fam_transect_grouped <- julymgpt1fam_transect %>%
  group_by(Transect, Family_grouped) %>%
  summarise(Percentage = sum(Percentage), .groups = "drop")

#consistent stacking order based on overall abundance!
family_order <- julymgpt1fam_transect_grouped %>%
  group_by(Family_grouped) %>%
  summarise(Total = sum(Percentage)) %>%
  arrange(Total) %>%
  pull(Family_grouped)

julymgpt1fam_transect_grouped$Family_grouped <- factor(
  julymgpt1fam_transect_grouped$Family_grouped,
  levels = family_order
)
#write.csv(julymgpt1fam_transect_grouped, "julymgpt1fam_transect_grouped.csv", row.names = FALSE)

```


Set up above (^) yields the following dataframes below (v) ~ dataframes below have been cleaned up and uploaded to github for faster reproduction)


read in the dataframes
```{r}
mgpt1fam_transect_grouped<- read.csv("mgpt1fam_transect_grouped.csv")
mgpt1fam_month_grouped<- read.csv("mgpt1fam_month_grouped.csv")
julymgpt1fam_transect_grouped <- read.csv ("julymgpt1fam_transect_grouped.csv")
mgp2fam_transect_grouped<- read.csv("mgpt2fam_transect_grouped.csv")

mgpt1fam_transect_grouped <- mgpt1fam_transect_grouped[, !names(mgpt1fam_transect_grouped) %in% "X"]
mgpt1fam_month_grouped    <- mgpt1fam_month_grouped[, !names(mgpt1fam_month_grouped) %in% "X"]
julymgpt1fam_transect_grouped <- julymgpt1fam_transect_grouped[, !names(julymgpt1fam_transect_grouped) %in% "X"]
mgp2fam_transect_grouped  <- mgp2fam_transect_grouped[, !names(mgp2fam_transect_grouped) %in% "X"]

View(mgpt1fam_transect_grouped)
View(mgpt1fam_month_grouped)
View(mgp2fam_transect_grouped)
View(julymgpt1fam_transect_grouped)
```

Scaling colors for visual comparisons 
(all the colors on the same scale) 
```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(RColorBrewer)

# Clean/standardize names
mgpt1fam_transect_grouped$Family_grouped <- trimws(as.character(mgpt1fam_transect_grouped$Family_grouped))
mgp2fam_transect_grouped$Family_grouped  <- trimws(as.character(mgp2fam_transect_grouped$Family_grouped))
mgpt1fam_month_grouped$Family_grouped    <- trimws(as.character(mgpt1fam_month_grouped$Family_grouped))
julymgpt1fam_transect_grouped$Family_grouped    <- trimws(as.character(julymgpt1fam_transect_grouped$Family_grouped))


#all families 
all_families <- unique(c(
  mgpt1fam_transect_grouped$Family_grouped,
  mgp2fam_transect_grouped$Family_grouped,
  mgpt1fam_month_grouped$Family_grouped,
  julymgpt1fam_transect_grouped$Family_grouped
))

# color palette *want contrast*
my_colors <- colorRampPalette(
  brewer.pal(12, "Paired")
)(length(all_families))

my_colors <- setNames(my_colors, all_families)

# color specific categories
my_colors[names(my_colors) %in% "Assigned at Higher Taxonomy"] <- "darkblue"
my_colors[names(my_colors) %in% "Rare Families"]              <- "grey80"
my_colors[names(my_colors) %in% "Unknown Family"]             <- "grey0"


# Set factor levels (for consistency)
mgpt1fam_transect_grouped$Family_grouped <- factor(
  mgpt1fam_transect_grouped$Family_grouped,
  levels = names(my_colors)
)

mgp2fam_transect_grouped$Family_grouped <- factor(
  mgp2fam_transect_grouped$Family_grouped,
  levels = names(my_colors)
)

mgpt1fam_month_grouped$Family_grouped <- factor(
  mgpt1fam_month_grouped$Family_grouped,
  levels = names(my_colors)
)
julymgpt1fam_transect_grouped$Family_grouped <- factor(
  julymgpt1fam_transect_grouped$Family_grouped,
  levels = names(my_colors)
)
```

Figure 1C ~ 2023 month stacked bar family plot
```{r}
#cleaning and ordering Monthc column
mgpt1fam_month_grouped$Month <- trimws(mgpt1fam_month_grouped$Month)
unique(mgpt1fam_month_grouped$Month)
mgpt1fam_month_grouped$Month <- factor(
  mgpt1fam_month_grouped$Month,
  levels = c( "June", "July", "December")
)

#plotting 2023 month
plot3 <- ggplot(mgpt1fam_month_grouped,
       aes(x = Month, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Month", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = my_colors)

plot3
```

Figure 1D (comparing july 2023 and july 2025)
```{r}
# Plot mgpt1 transect
plot1<- ggplot(julymgpt1fam_transect_grouped,
       aes(x = Transect, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Transect", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = my_colors)
plot1
# Plot mgpt2 transect
plot2<- ggplot(mgp2fam_transect_grouped,
       aes(x = Transect, y = Percentage, fill = Family_grouped)) +
  geom_bar(stat = "identity") +
  labs(x = "Transect", y = "Relative Abundance (%)", fill = "Family") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = my_colors)
plot2
```


combining both plots and putting key in alphabetical order! 

```{r}

library(cowplot)
library(ggplot2)

# -----------------------------
# Alphabetical order of ALL families
# -----------------------------

family_order <- sort(unique(c(
  as.character(julymgpt1fam_transect_grouped$Family_grouped),
  as.character(mgp2fam_transect_grouped$Family_grouped)
)))

# Make sure both datasets use the same factor levels
julymgpt1fam_transect_grouped$Family_grouped <- factor(
  julymgpt1fam_transect_grouped$Family_grouped,
  levels = family_order
)

mgp2fam_transect_grouped$Family_grouped <- factor(
  mgp2fam_transect_grouped$Family_grouped,
  levels = family_order
)


# -----------------------------
# July 2023 plot
# -----------------------------

plot1 <- ggplot(
  julymgpt1fam_transect_grouped,
  aes(x = Transect, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Transect",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )


# -----------------------------
# July 2025 plot
# -----------------------------

plot2 <- ggplot(
  mgp2fam_transect_grouped,
  aes(x = Transect, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Transect",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )


# -----------------------------
# Create combined legend
# -----------------------------

legend_df <- data.frame(
  Family_grouped = factor(
    family_order,
    levels = family_order
  ),
  Percentage = 1,
  Transect = "Legend"
)

legend_plot <- ggplot(
  legend_df,
  aes(
    x = Transect,
    y = Percentage,
    fill = Family_grouped
  )
) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  labs(fill = "Family") +
  theme_bw() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.1, "cm")
  )


# Extract combined alphabetical legend
legend <- cowplot::get_legend(legend_plot)


# -----------------------------
# Combine plots + legend
# -----------------------------

combined.2d <- cowplot::plot_grid(
  plot1,
  plot2,
  legend,
  ncol = 3,
  rel_widths = c(1, 1, 0.5)
)

combined.2d
```



cleaned-up combo plot
```{r}
library(cowplot)
library(ggplot2)
library(grid)

# -----------------------------
# Alphabetical order of ALL families
# -----------------------------

family_order <- sort(unique(c(
  as.character(julymgpt1fam_transect_grouped$Family_grouped),
  as.character(mgp2fam_transect_grouped$Family_grouped)
)))

# Make sure both datasets use the same factor levels
julymgpt1fam_transect_grouped$Family_grouped <- factor(
  julymgpt1fam_transect_grouped$Family_grouped,
  levels = family_order
)

mgp2fam_transect_grouped$Family_grouped <- factor(
  mgp2fam_transect_grouped$Family_grouped,
  levels = family_order
)


# -----------------------------
# July 2023 plot
# -----------------------------

plot1 <- ggplot(
  julymgpt1fam_transect_grouped,
  aes(x = Transect, y = Percentage, fill = Family_grouped)
) +
  geom_bar(
    stat = "identity",
    width = 0.6
  ) +
  labs(
    x = "Transect",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )


# -----------------------------
# July 2025 plot
# -----------------------------

plot2 <- ggplot(
  mgp2fam_transect_grouped,
  aes(x = Transect, y = Percentage, fill = Family_grouped)
) +
  geom_bar(
    stat = "identity",
    width = 0.6
  ) +
  labs(
    x = "Transect",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )


# -----------------------------
# Create compact legend
# -----------------------------

legend_df <- data.frame(
  Family_grouped = factor(
    family_order,
    levels = family_order
  ),
  Percentage = 1,
  Transect = "Legend"
)

legend_plot <- ggplot(
  legend_df,
  aes(
    x = Transect,
    y = Percentage,
    fill = Family_grouped
  )
) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  labs(fill = "Family") +
  theme_bw() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.spacing.y = unit(0.05, "cm"),
    legend.margin = margin(0, 0, 0, 0),
    legend.box.margin = margin(0, 0, 0, 0),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

# Extract legend
legend <- cowplot::get_legend(legend_plot)


# -----------------------------
# Combine plots + legend
# -----------------------------

plots_row <- cowplot::plot_grid(
  plot1,
  plot2,
  ncol = 2,
  rel_widths = c(1, 1),
  align = "h"
)

combined.2d_clean <- cowplot::plot_grid(
  legend,
  plots_row,
  ncol = 2,
  rel_widths = c(0.3, 1),
  align = "h"
)

combined.2d_clean
```
update:
```{r}
# -----------------------------
# Create compact vertical legend
# -----------------------------

legend_df <- data.frame(
  Family_grouped = factor(
    family_order,
    levels = family_order
  ),
  Percentage = 1,
  Transect = "Legend"
)

legend_plot <- ggplot(
  legend_df,
  aes(
    x = Transect,
    y = Percentage,
    fill = Family_grouped
  )
) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order,
    drop = FALSE
  ) +
  labs(fill = "Family") +
  guides(
    fill = guide_legend(
      ncol = 1,
      byrow = TRUE
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.spacing.y = unit(0.05, "cm"),
    legend.margin = margin(0, 0, 0, 0),
    legend.box.margin = margin(0, 0, 0, 0),
    panel.grid = element_blank(),
    panel.border = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

legend <- cowplot::get_legend(legend_plot)
```


```{r}
combined.2d_clean <- cowplot::plot_grid(
  legend,
  plots_row,
  ncol = 2,
  rel_widths = c(0.3, 1),
  align = "h"
)

combined.2d_clean
```



redoing month plot to match!
```{r}
# -----------------------------
# Figure 1C: 2023 month stacked bar family plot
# -----------------------------

# Clean Month column
mgpt1fam_month_grouped$Month <- trimws(mgpt1fam_month_grouped$Month)

# Set month order
mgpt1fam_month_grouped$Month <- factor(
  mgpt1fam_month_grouped$Month,
  levels = c("June", "July", "December")
)

# Set family order alphabetically
family_order <- sort(names(my_colors))

mgpt1fam_month_grouped$Family_grouped <- factor(
  mgpt1fam_month_grouped$Family_grouped,
  levels = family_order
)

# Create plot
plot3 <- ggplot(
  mgpt1fam_month_grouped,
  aes(x = Month, y = Percentage, fill = Family_grouped)
) +
  geom_bar(stat = "identity") +
  labs(
    x = "Month",
    y = "Relative Abundance (%)",
    fill = "Family"
  ) +
  theme_bw() +
  theme(
    legend.text = element_text(size = 8),      # family names
    legend.title = element_text(size = 9),    # "Family"
    legend.key.size = unit(0.4, "cm"),         # colored boxes
    legend.spacing.y = unit(0.1, "cm"),        # spacing between items
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_manual(
    values = my_colors,
    breaks = family_order
  )

plot3
```

~~~~~~~~~~~~~~~~~~~~~~~~~~PAM Stats (both 2023 & 2025)~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
```{r}
mgpamupdate <- read.csv("MGPAM_UPDATED.csv")
```

2way anova
```{r}
mgp_model <- aov(Y.II. ~ MatLoc * Month_Year, data = mgpamupdate)
summary(mgp_model)
```

```{r}
TukeyHSD(mgp_model)
```
*stat difference between months (Month_Year, p = 0.0199 )
*tukey test finds that specific difference is between: (Dec_2023-July_2023, p = 0.0129673)

lets check assumptions: ~ passes (residual vs fitted scatter looks ehh) but overall model is good to use
```{r}
#normality
plot(mgp_model, which = 1)   # residuals vs fitted
plot(mgp_model, which = 2)   # QQ plot
shapiro.test(residuals(mgp_model))
#homogeneity
car::leveneTest(Y.II. ~ MatLoc * Month_Year, data = mgpamupdate)
# Visual check: residuals vs fitted
plot(fitted(mgp_model), residuals(mgp_model),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")
# Look for random scatter around zero
```

~~~~Code for Final Combination Plot!

clean up plots


```{r}
# -----------------------------
# Make bar plots slightly narrower
# and keep legend on the left
# -----------------------------

combined.2d <- cowplot::plot_grid(
  legend,
  cowplot::plot_grid(
    plot1,
    plot2,
    ncol = 2,
    rel_widths = c(1, 1)
  ),
  ncol = 2,
  rel_widths = c(0.40, 1),
  align = "h"
)

combined.2d
```

```{r}
# Remove gridlines from both plots

plot3_clean <- plot3 +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
     panel.border = element_blank()
  )

plot3_clean
```
```{r}
heatmap_topfamily2_clean <- heatmap_topfamily2 +
  theme(
    panel.border = element_blank()
  )

heatmap_topfamily2_clean
```

dowload each figure - journal recommendation as 4 stand alone figures (going to do 3 figures - since final panel goes together)
```{r}
ggsave(
  filename = "monthbarplot.pdf",
  plot = plot3_clean,
  width = 5,
  height = 5,
  units = "in"
)
```

```{r}
ggsave(
  filename = "heatmap.pdf",
  plot = heatmap_topfamily2_clean,
  width = 5,
  height = 5,
  units = "in"
)
```

```{r}
ggsave(
  filename = "combined.2d_clean.pdf",
  plot = combined.2d_clean,
  width = 7.09,
  height = 8.66,
  units = "in"
)
```

