---
title: "MG_REPROCESS"
output: html_document
date: "2024-05-13"
---
Helpful Terminology:

OTUs (Operational Taxonomic Units):OTUs are clusters of similar sequences grouped together based on a defined similarity threshold, typically used to represent taxa in microbial ecology.Each OTU corresponds to a specific taxonomic unit, which could be at any taxonomic level (species, genus, etc.).

ASVs (Amplicon Sequence Variants) ASVs represent unique sequences in your dataset without clustering. Each ASV is a single, unique DNA sequence.Unlike OTUs, ASVs do not rely on a similarity threshold for clustering. Instead, they use exact sequence matching. (however these terms are typically used interchangeable and from my understanding ASV is "better")

Reads:Reads are sequences obtained from your metabarcoding experiment.
The abundance of reads for each OTU in a sample indicates how many times sequences belonging to that OTU (or ASV) were observed in that sample.

Taxa2?

helpful link! https://benjjneb.github.io/dada2/tutorial.html
installing required packages
```{r}
# Installing/loading packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packags("BiocManager")

# Installing dada2 package from CRAN
install.packages("dada2", dependencies = TRUE)
```

```{r}
# Loading dada2 package
library(dada2)

# Checking the installed version of dada2
packageVersion("dada2")
```

```{r}
install.packages("ShortRead")
```

```{r}
library(ShortRead)
```

```{r}
library(dada2)
library(Rcpp)
```

1. STARTED A NEW PROJECT 
```{r}
# Checking Current Working Directory
print(getwd())

# Defining Path Relative to Current Directory
path <- "FINALFASTQMG/" 

# Checking if Path Exists
if (!file.exists(path)) {
  # Creating Directory if it doesn't exist
  dir.create(path)
  print(paste("Created directory:", path))
} else {
  print(paste("Directory already exists:", path))
}
```

2. Generates sorted lists of filenames - lists them by R1 & R2
path: the directory containing the file path
pattern: selecting files that contain that in their name
sort: putting these files in alphabetically and stored in fnFs & fnRs
```{r}
fnFs <- sort(list.files(path, pattern = "_R1_16S_final.fastq", full.names = TRUE))

fnRs <- sort(list.files(path, pattern = "_R2_16S_final.fastq", full.names = TRUE))
```

3. This is used to generate the list of sample names by extracting the 3rd part of the file name (this is the part that should be containing the sample name)
```{r}
get.sample.name <- function(fname) strsplit(basename(fname), "_")[[1]][3]
sample.names <- unname(sapply(fnFs, get.sample.name))
```

```{r}
print(sample.names)
```

4. TO VIEW new fnFs (forward) & fnRs (reverse) FILES - want to make sure you have all the samples
```{r}
print(fnFs)
```
```{r}
print(fnRs)
```

5. identifying primers and all orientations of primers 
```{r}
FWD <- "GTGYCAGCMGCCGCGGTAA"  
REV <- "GGACTACNVGGGTWTCTAAT"  
```

This is creating all orientations of the primers
```{r}
allOrients <- function(primer) {
  require(Biostrings)
  dna <- DNAString(primer)  # The Biostrings works w/ DNAString objects rather than character vectors
  orients <- c(Forward = dna, Complement = complement(dna), Reverse = reverse(dna), 
               RevComp = reverseComplement(dna))
  return(sapply(orients, toString))  # Convert back to character vector
}
FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)
FWD.orients
REV.orients

```

6. Filters and trims the original fastq files (fnFs and fnRs), removes reads with too many ambiguous bases (N's), and saves the filtered files in a subdirectory named "filtN" within the original directory. --so my filtN will be within FINALFASTQ MG_WD folder 

*it takes a long time ~ about 30 mins for 182 samples - can see files being created in the FiltN folder
```{r}
fnFs.filtN <- file.path(path, "filtN", basename(fnFs)) # Put N-filtered files in filtN/ subdirectory
fnRs.filtN <- file.path(path, "filtN", basename(fnRs))
filterAndTrim(fnFs, fnFs.filtN, fnRs, fnRs.filtN, maxN = 0, multithread = TRUE, compress=TRUE)
```


7. Double check: Looking for hits of primers in any orientation
*we want this to be 0 ~ if not then we cut the sequences with the primers!
yay no primers!!
```{r}
primerHits <- function(primer, fn) {
  # Counts number of reads in which the primer is found
  nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
  return(sum(nhits > 0))
}

rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn = fnFs.filtN[[4]]), 
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn = fnRs.filtN[[4]]), 
      REV.ForwardReads = sapply(REV.orients, primerHits, fn = fnFs.filtN[[4]]), 
      REV.ReverseReads = sapply(REV.orients, primerHits, fn = fnRs.filtN[[4]]))
```

8. TIME TO VISUALIZE THE RAW DATA!! QC plots
a. Looking at quality profile of ALL R1 reads***
You are looking at where the line on the graphs begin to plummet (this is where the quality of the reads goes down) we want to cut prior to where the quality goes down. You should look at every single sample.
```{r}
plotQualityProfile(fnFs.filtN[c(1:5)]) 
```
A1, A10, A11, A12, A2 

```{r}
plotQualityProfile(fnFs.filtN[c(6:10)])
```
A3, A4, A5, A6, A7

```{r}
plotQualityProfile(fnFs.filtN[c(11:16)]) 
```
A8, A9, B1, B10, B11, B12

```{r}
plotQualityProfile(fnFs.filtN[c(17:22)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(23:28)]) 
```


```{r}
plotQualityProfile(fnFs.filtN[c(29:34)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(35:40)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(41:50)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(51:60)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(61:70)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(71:80)]) 
```

```{r}
plotQualityProfile(fnFs.filtN[c(81:91)]) 
```
*All reads look good for a cut-off around 210


b. NOW - Looking at quality profile of R2 reads***

```{r}
plotQualityProfile(fnRs.filtN[c(1:5)]) 
```
A1, A10, A11, A12, A2 

```{r}
plotQualityProfile(fnRs.filtN[c(6:10)])
```
A3, A4, A5, A6, A7

```{r}
plotQualityProfile(fnRs.filtN[c(11:16)]) 
```
A8, A9, B1, B10, B11, B12

```{r}
plotQualityProfile(fnRs.filtN[c(17:22)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(23:28)]) 
```


```{r}
plotQualityProfile(fnRs.filtN[c(29:34)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(35:40)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(41:50)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(51:60)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(61:70)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(71:80)]) 
```

```{r}
plotQualityProfile(fnRs.filtN[c(81:91)]) 
```
*These reads look great! Will truncate at 210 for both R1 & R2:)

9. FILTER AND TRIM:

This next step takes a while ...  1 hour
This is trimming based on quality profile plots cut off & uploading new, trimmed files into a folder (called trimmed) - to check if code is working look for files popping up in the new trimmed folder 

truncLen=c(210,210), #specifies where to be truncated - leaves plenty of overlap
maxN=0, #DADA does not allow ambiguous(N) bases, so 0 
maxEE=c(2,2), #allow 2 expected errors in a read (for both forward & reverse reads)
truncQ=2, #trims low quality bases from the ends of their reads (2 is threshold)
rm.phix=TRUE, #remove reads matching phiX genome
matchIDs=TRUE, #enforce matching between id-line sequence identifiers of F and R reads
compress=TRUE, multithread=TRUE) #On Windows set multithread=FALSE

```{r}
filt_path <- file.path(path, "trimmed")
if(!file_test("-d", filt_path)) dir.create(filt_path)

filtFs <- file.path(filt_path, paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(filt_path, paste0(sample.names, "_R_filt.fastq.gz"))

#maxEE = 2,2 - allows 2 expected errors
#Trunclen = 175 - here I decided to do 200 because they all looked so good except maybe 13 which even looked pretty good still
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, 
                     truncLen=c(210,210), #leaves plenty of overlap
                     maxN=0, #DADA does not allow Ns
                     maxEE=c(2,2), #allow 1 expected errors, where EE = sum(10^(-Q/10)); more conservative, model converges
                     truncQ=2, 
                     #trimLeft=c(18,20), #N nucleotides to remove from the start of each read
                     rm.phix=TRUE, #remove reads matching phiX genome
                     matchIDs=TRUE, #enforce matching between id-line sequence identifiers of F and R reads
                     compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE

head(out)
tail(out)
write.csv(out,file="out.csv",row.names=TRUE,quote=FALSE)
```

10. LEARNING ERROR RATES NEXT 
Error rates typically range from .1%-1% per base pair sequence 
Types of potential errors: PCR errors, sequencing errors, chimeric reads, contaminants
This step is checking for the rates of errors in the reads
It sets an option in the DADA2 pipeline to allow up to 30 iterations during the error rate consistency estimation step
```{r}
setDadaOpt(MAX_CONSIST=30) #increase number of cycles to allow convergence
```

This will take 20 mins
```{r}
errF <- learnErrors(filtFs, multithread=TRUE) 
```

This will also take 20 mins
```{r}
errR <- learnErrors(filtRs, multithread=TRUE)
```

11. Visualizing estimated error rates
- error rates should decline with increasing qual score
- the red line is based on definition of quality score alone
- theblack line is estimated error rate after convergence
- dots are observed error rate for each quality score
- you want the black dots to roughly fall along the black line 
```{r}
plotErrors(errF, nominalQ=TRUE) 
plotErrors(errR, nominalQ=TRUE) 
```
*Everything looks reasonable (no strong deviations from black lines and black points), so we proceed with confidence

12. Deplicate Reads: 

-  Dereplication in bioinformatics means grouping identical DNA sequences together and counting how many times each sequence appears. This helps speed up the analysis by avoiding repeating the same calculations for identical sequences : For example, if you have 100 copies of the same sequence, instead of treating each copy separately, dereplication counts it as one sequence with an abundance of 100.

- In DADA2 it keeps track of the quality (or reliability) of each unique sequence. It calculates an average quality score for each sequence by considering the quality scores of all the copies. 

-This helps DADA2 make fewer mistakes when it tries to figure out the true sequences from the noisy data generated by sequencing machines.
*DADA2's denoising step can more effectively distinguish between true biological variants and sequencing artifacts

- In simple terms, dereplication helps bioinformatics tools work faster by grouping identical sequences together, and DADA2 uses quality information to make better guesses about the real DNA sequences

This takes 5 mins
```{r}
derepFs <- derepFastq(filtFs, verbose=TRUE)
derepRs <- derepFastq(filtRs, verbose=TRUE)
# Name the derep-class objects by the sample names - now the dereps are named A1-H7
names(derepFs) <- sample.names
names(derepRs) <- sample.names
```
13. Sample Inference: 
-We are now ready to apply the core sample inference algorithm to the filtered and trimmed de-replicate reads, using the error rates identified above
- essentially, this step is denoising the depreplicated reads using error rates (cleaning up the unique DNA sequences we found, using information about how often sequencing mistakes happen, to get a more accurate picture of the real DNA sequences)
-dadaFs & dadaRs now stores the denoised, dereplicated sequences that are now cleaned up - along with also holding their respective abundances and quality scores (produced as a result of the algorithm)
*this takes a while....like 2 hours. you can tell its running if you see samples popping up below. 
```{r}
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)
dadaRs <- dada(derepRs, err=errR, multithread=TRUE)
```
Inspecting the Returned dada-class object
```{r}
dadaFs[[1]]
dadaRs[[1]]
```
- Now, looking at the dada-class objects by sample
- Will tell how many true (biologically-relevant) sequence variants from the unique sequences in the first sample for both the forward and reverse reads
- Essentially, this tells us how much it denoised, filtered, and error-corrected the data - to make cleaned up reads (sequence variants) that are more biologically relevant!

14. Merge Paired Reads 
- We now merge the forward and reverse reads together to obtain the full denoised sequences.
- Paired reads that do not exactly overlap are removed
```{r}
mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, minOverlap = 12,
  maxMismatch = 0,
  returnRejects = FALSE,
  propagateCol = character(0),
  justConcatenate = FALSE,
  trimOverhang = FALSE,
  verbose = TRUE)

#Inspect the merger data.frame from the first sample
head(mergers[[1]])
```
```{r}
summary((mergers[[1]]))
```
We now have a data.frame for each sample with the merged sequence, its abundance, and the indices of the merged forward and reverse denoised sequences. Paired reads that did not exactly overlap were removed by mergePairs.:)

More Background on derep and dada: 
DEREP Description: A list with the following three members, uniques: Named integer vector. Named by the unique sequence, valued by abundance, quals: Numeric matrix of average quality scores by position for each unique. Uniques are rows, positions are cols., map: Integer vector of length the number of reads, and value the index (in $uniques) of the unique to which that read was assigned.

DADA Description: A multi-item List with the following named values...denoised: Integer vector, named by sequence valued by abundance, of the denoised sequences; clustering  An informative data.frame containing information on each cluster; sequence: A character vector of each denoised sequence. Identical to names(denoised).; quality: The average quality scores for each cluster (row) by position (col); map: Integer vector that maps the unique (index of derep unique) to the denoised sequence (index of dada denoised); birth_subs: A data.frame containing the substitutions at the birth of each new cluster; trans: The matrix of transitions by type (row), eg. A2A, A2C..., and quality score (col) observed in the final output of the dada algorithm; err_in: The err matrix used for this invocation of dada.; err_out: The err matrix estimated from the output of dada. NULL if err_function not provided.; opts: A list of the dada_opts used for this invocation of dada.


15. Construct Sequence table!!

a higher-resolution version of the “OTU table” produced by classical methods. 
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
#91 samples 19999 sequences
saveRDS(seqtab, file="seqtab_mg_16S.rds")
write.csv(seqtab, file="seqtab_mg_16S.csv")

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
plot(table(nchar(getSequences(seqtab)))) #real variants appear to be right in that 240-270 window
```
The sequence table is a matrix with rows corresponding to (and named by) the samples, and columns corresponding to (and named by) the sequence variants.This table contains 19999 ASVs and the lengths of our merged sequences fall within expected range for V4 amplican (240-270)

```{r}
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% seq(240,270)] #again, being fairly conservative with length
saveRDS(seqtab2, file="seqtab2_mg_16S.rds")
write.csv(seqtab2, file="seqtab2_mg_16S.csv")
```
Sequences that are much longer or shorter than expected may be the result of non-specific priming, and may be worth removing. This is removing non-target-length sequences from my sequence table 

16. Removing chimeras!!

Chimeras are hybrid sequences formed when 2 or more different original sequences are fused. These are errors that must be removed from the reads. The core dada method removes substitution and indel errors, but chimeras remain. Fortunately, the accuracy of the sequences after denoising makes identifying chimeras easier than it is when dealing with fuzzy OTUs: chimeric sequences are identified if they can be reconstructed by combining left segment and right segment from 2 or more parent sequences. 

This step takes 5 mins. Removing chimeras and saving this new file 
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim) 
saveRDS(seqtab.nochim, file="seqtab.nochim_mg_16S.rds")
write.csv(seqtab.nochim, file="seqtab.nochim_mg_16S.csv")

sum(seqtab.nochim)/sum(seqtab2)
```
A bimera = a 2 parent chimera. Essentially, this is telling us that 97% of the sequences were fine, so will continue!

17. Track Sequence Reads 
A final check of progressive, this will look at the number of reads that made it through each step in the pipeline. Sanity check! 
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(mergers, getN), rowSums(seqtab2), rowSums(seqtab.nochim))
colnames(track) <- c("input", "filtered", "denoised", "merged", "tabled", "nonchim")
rownames(track) <- sample.names
head(track)
tail(track)

write.csv(track,file="MG_2024_16S_readstats.csv",row.names=TRUE,quote=FALSE)

```
You are checking to make sure majority of raw reads were kept and that there is no over-large drop associated with any single step. (H2-H6 is looking a little sketchy and H7 is water, so why so many reads still?) The reads in H7 will have to be removed from all other samples.

18. Assign Taxonomy 

This will take a long time. 30 minutes - 1 hour. Should be in a new file labelled "MG_2024_16s_taxa"
assignTaxonomy: function in dada2 package. Assigns taxonomy of the seqtab.nochim matrix
Multithread: how many thought processes can occur at the same time 
```{r}
taxa <- assignTaxonomy(seqtab.nochim, "~/Desktop/ENVSTATS_LAB/FINALFASTQMG/silva_nr99_v138.1_train_set.fa", multithread=TRUE)
saveRDS(taxa, file="MG_2024_16S_taxa.rds") 
```

Reading in taxa datafile. 
```{r}
taxa <- readRDS("MG_2024_16S_taxa.rds")
```

This is showing taxonomy identified 
```{r}
unname(head(taxa))
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
taxa <- readRDS("MG_2024_16S_taxa.rds")
```

This next step uses the "addSpecies" function in dada2. It takes as input the output from assignTaxonomy and a separate reference database specifically designed for species-level classification. It attempts to match the sequences to the reference species database, providing an additional species-level classification where possible. *addSpecies function will not work on this computer -- need Schyler or Maya to do it. 

Run this on ondemand because taking up too much memory 
https://ondemand.rc.unc.edu/pun/sys/dashboard/batch_connect/sys/rstudio-r/session_contexts/new

Go to apps, click r studio:
download taxa file onto proj/mccoy/maddy 
download silva file -- REZIP ON TERMINAL FIRST using g zip command
-Navigate to desktop (cd) & (ls)
-Navigate to correct folder 
-use g zip to zip silva file
-exit terminal 

open up r studio server:
*launch and upload the code below onto this - this is so that there is enough memory on longleaf to run this, since there is not enough on r
*another option if you want to stay on this computer and attempt to run, is to use the n=2000 function - this will run in chunks so that it doesn't overload memory - I am NOW attempting this since my r kept crashing on the on demand longleaf site

***This step did not run for me 
- I attempted both strategies multiple times, so I had Schyler do it on the ondemand longleaf
- It took him 3 minutes on ondemand long leaf. Under "additional job submission arguments" he changed it to "--mem=14G" and he allowed for 3 hours. Screenshots of code can be found in computer/16s folder
```{r}
taxa.plus <- addSpecies(taxa, "~/Desktop/ENVSTATS_LAB/FINALFASTQMG/silva_species_assignment_v138.1_1.fa.gz",tryRC=TRUE,verbose=TRUE)
saveRDS(taxa.plus, file="MG_2024_16S_taxa.plus.rds")
```
*Note since reading in Schyler's, was saved as "taxa.plus.rds" instead of "MG_2024_16S_taxa.plus.rds"

Loading in taxa.plus file
```{r}
taxa.plus <- readRDS("taxa.plus.rds")
View(taxa.plus)
```

Making taxa.plus & taxa files into csvs.
```{r}
write.csv(taxa.plus, file="taxa.plus.csv")
write.csv(taxa, file="MG_2024_16S_taxa.csv")
```

Reading in Previously saved datafiles
seqtab.nochim: the is a sequence table (rows are samples and columns are asvs) 
taxa: this is assigned taxonomy to the asvs 
taxa.plus: this is more specific assigning of taxonomy (species level) of asvs
```{r}
seqtab.nochim <- readRDS("seqtab.nochim_MG_16S.rds")
taxa <- readRDS("MG_2024_16S_taxa.rds")
taxa.plus <- readRDS("taxa.plus.rds")
```


19. handoff 2 phyloseq 
Phyloseq is a package in r that helps with  import, storage, analysis, and graphical display of phylogenetic sequencing data. 

Installing packages
```{r}
BiocManager::install("phyloseq")
install.packages("Rmisc")
```

library 
```{r}
library('phyloseq')
library('ggplot2')
library(cowplot)
library(ShortRead)
library('Rmisc')
```

Reading in metadata csv file with additional info about samples. Make sure to add in metafile whatever you want to compare your microbiome data to (for me it is transect location)
```{r}
samples.out <- rownames(seqtab.nochim)
samdf<-read.csv("MG_metadata.csv")
head(samdf)
rownames(samdf) <- samdf$id
```
```{r}
View(samdf)
row.names(samdf) <- samdf$Sample
```

This is combining seqtab.nochim with taxa.plus and samdf
- ensure row names match!
```{r}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa.plus)) 
ps
```
Results from ps are important to take note of: *Mine listed below

phyloseq-class experiment-level object
otu_table()   OTU Table:         [ 18820 taxa and 91 samples ]
sample_data() Sample Data:       [ 91 samples by 5 sample variables ]
tax_table()   Taxonomy Table:    [ 18820 taxa by 7 taxonomic ranks ]

TAX.GLOM function is aggregating similar ASV into taxonomic ranks (phylum, class, order, family, genus). Since I am interested in comparing microbiomes between transect location then I change my x to "transect" from my metadata. 

Multiple reads can match an ASV - this is how you get abundance data. (ASV is the specific taxa being identified)
```{r}
ps_glom <- tax_glom(ps, "Family")
```

```{r}
plot_bar(ps_glom, x="Transect", fill="Family")+
  theme(legend.position="none")
```
This is showing you how many reads at each transect location. N/A is negative control. Should be smallest. 

Making the ASV names shorter
```{r}
ids <- paste0("sq", seq(1, length(colnames(seqtab.nochim))))
```

Making output fasta file - this will be blasted for taxa in a bit! - IF NOT doing BLAST, don't bother teehee
```{r}
path='~/Desktop/ENVSTATS_LAB/FINALFASTAFILE.fasta'
uniquesToFasta(seqtab.nochim, path, ids = ids, mode = "w", width = 20000)
```

BLAST could be completed next. This is to remove Eukaryotes. (AN extra step de-contamination step). Ondemand site to do the BLAST. We are not doing the BLAST because we are working specifically with bacterial primers. This is not a standard step, just an extra one if you really are scared of eukaryotes in your data. We will pull out euks from silva later, so not super important. Will skip blast step for now. if you want to do it, see Maya's code. 

20. Removing Contaminants. 

A) First Renaming Stuff:
This assigns the values in 'ids' to be the column names of the 'seqtab.nochim' table. Creating taxa2 by combining taxa.plus with its row names. Now assigning values in ids to be the names of the taxa2 table. 
Essentially, this is renaming stuff.
```{r}
colnames(seqtab.nochim)<-ids
taxa2 <- cbind(taxa.plus, rownames(taxa.plus)) #retaining raw sequence info before renaming
rownames(taxa2)<-ids
```

This code constructs a phyloseq object named ps by combining the ASV table "seqtab.nochim", sample metadata "samdf", and NEW taxonomy table "taxa2" 
```{r}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa2))
ps
```
ps output:

phyloseq-class experiment-level object
otu_table()   OTU Table:         [ 18820 taxa and 91 samples ]
sample_data() Sample Data:       [ 91 samples by 5 sample variables ]
tax_table()   Taxonomy Table:    [ 18820 taxa by 8 taxonomic ranks ]


Saving new (still contaminated) ps file, in case r crashes: 
```{r}
saveRDS(ps, file = "Mg_ps.rds")
```


B) Now, removing mitochondria, chloroplasts & non-bacteria taxa.

Removing mitochondria!!
```{r}
ps.mito <- subset_taxa(ps, (Family=="Mitochondria"))
ps.mito
```
314 mito-sequences present - these will need to be removed. 

```{r}
ps.chlor <- subset_taxa(ps, (Order=="Chloroplast"))
ps.chlor
```
102 chloro-sequences present - these will need to be removed also. 

```{r}
ps.notbact <- subset_taxa(ps, (Kingdom!="Bacteria") | is.na(Kingdom))
ps.notbact
```
2115 notbacteria sequences present - these will ALSO need to be removed. 

```{r}
ps.nomito <- subset_taxa(ps, (Family!="Mitochondria") | is.na(Family))
ps.nomito
```
This is data with mitochondria removed (total ps sequences - ps.mito sequences)

```{r}
ps.nochlor <- subset_taxa(ps.nomito, (Order!="Chloroplast") | is.na(Order))
ps.nochlor
```
This is data with mitochondria AND chlorophyll removed (ps no.mito - ps chlor)

```{r}
ps.clean <- subset_taxa(ps.nochlor, (Kingdom=="Bacteria"))
ps.clean
saveRDS(ps, file = "Mg_psclean.rds")
```
This is clean data without mitochondria AND chlorophyll AND non -bacteria members removed (ps no.chloro - ps.notbacter)

C) Just for funsies - This is how much archaea was identified 
```{r}
ps.arch <- subset_taxa(ps.nomito, (Kingdom=="Archaea"))
ps.arch 
```
ps.arch output!!
phyloseq-class experiment-level object
otu_table()   OTU Table:         [ 2106 taxa and 91 samples ]
sample_data() Sample Data:       [ 91 samples by 5 sample variables ]
tax_table()   Taxonomy Table:    [ 2106 taxa by 8 taxonomic ranks ]


D) Removing negative controls from samples!

Installing decontam package: 
```{r}
BiocManager::install("decontam")
```

```{r}
library(decontam)
```

This code first converts sample data (ps.clean) to dataframe. 
Then it calculates the library size (total number of reads) for each sample in ps.clean and adds a new column 'library size' into the dataframe 'df'
Then df is ordered in asceding order by library size
The df$index creates a new column 'index' that contains sequential numbers from 1 to the number of rows in the dataframe (x axis in plot)
ggplot: creats scatterplot. Index is the x axis, library size is the y axis, and colors are the points by the transect variable (this is because this is the variable we are comparing microbiome data to eachother)

```{r}
df <- as.data.frame(sample_data(ps.clean)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps.clean)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Transect)) + 
  geom_point()
```
N/A is neg control so should be clumped at the bottom (pink points). Looking for trends in # reads compared to transect location.

```{r}
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Month)) + 
  geom_point()
```
N/A is neg control so should be clumped at the bottom (purple points). Looking for trends in # reads compared to month


first line creates a new column 'is.neg' in the object of 'ps.clean'.checks to make sure the 'Sample' column in the sample data is equal to 'H7' - this is the negative control 
next line uses "isContaminant" function to identify potential contaminants in ps.clean - puts .5 as a threshold (if a feature has a 50% or more probability of being a contaminant, it is removed)
last line creates a table of the contaminant column in the new contamdf.prev dataframe
```{r}
sample_data(ps.clean)$is.neg <- sample_data(ps.clean)$Sample == "H7"
contamdf.prev <- isContaminant(ps.clean, neg="is.neg",threshold=0.5)
table(contamdf.prev$contaminant)
```

FALSE  TRUE 
16223    66  
This means out of my 16,289 OTUS/ASVs, 66 were identified as contaminants and 16,223 were not. The identification of contaminants is throughout all samples (using the negative control to help identify likely contaminants)


The first line applies a transformation to the sample counts in the ps.clean (either presence or absent)
identifying the neg controls in the ps.pa object 
identifying the samples that are not negative controls 
Summary: transforming to presence/absence data and subsetting samples that are neg controls vs all other samples (only neg control samples are Sample==H7)
```{r}
ps.pa <- transform_sample_counts(ps.clean, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample == "H7", ps.pa) 
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample != "H7", ps.pa)
```

Make data.frame of prevalence in positive and negative samples
```{r}
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa),
                    contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```
This is visualizing the contaminants within all the samples. (contaminants identified using negative control)

This is contaminant taxa (identified above) from all your samples (from the phyloseq object)
```{r}
ps.clean1 <- prune_taxa(!contamdf.prev$contaminant,ps.clean)
```

Now, you are removing your negative controls as well - because we do not need these anymore
```{r}
ps.cleaner <- subset_samples(ps.clean1,(Sample!="H7"))
ps.cleaner
```
*SINCE H7 was my only neg control, should have one less # of samples (so 90 samples)
16,223 "non-contaminated" taxa and 90 samples 


Saving files: *PS.CLEANER IS MOST IMPORTANT AT THIS POINT. This should have clean data (neg controls and contaminant sequences from neg controls removed in all samples)


```{r}
saveRDS(ps.cleaner, file ="MG_16S_ps.cleaner.RDS")
saveRDS(taxa, file ="MG_16S_taxa.Rdata")
saveRDS(taxa.plus, file ="MG_16S_taxa.plus.Rdata")
saveRDS(taxa2, file ="MG_16S_taxa2.Rdata")
ps.cleaner <- readRDS("MG_16S_ps.cleaner.RDS")
```

What are these files? 
Taxa: taxonomy assigned to sequence table 
Taxa.plus: taxonomy assigned to sequence table (more specific, species level identification included)
ps.cleaner: phyloseq object with contaminants (identified from neg control) removed as well as the negative control itself removed 



21. Rarefying steps!
Rarefying is essentially throwing away data based on choosing the lowest sequencing depth and ensuring all samples have the same sequencing depth. 
```{r}
library(vegan)
library(permute)
```

This is turning the cleaned asv data (ps.cleaner) along with the metadata into an otu table, and then a dataframe. This is to help with future r manipulations and steps.
```{r}
seqtab.cleaner <- data.frame(ps.cleaner@otu_table)
samdf.cleaner <- data.frame(ps.cleaner@sam_data)
```
 
Making a rarefaction curve.
```{r}
rarecurve(seqtab.cleaner,step=1000,label=FALSE)
```
This is after removing contaminants based on SILVA (only identification) & negative controls (removing taxa in all samples based on negative controls) 
This is looking at a rarefaction curve, which analyzes species richness of your samples at different sequencing depths. 

Sequencing depth is the total number of sequencing reads obtained for each sample. You want sequencing depths to be similar for each sample. 

Rarefaction curves plot the number of observed species (or ASVs/OTUs) against the number of sequences sampled. Here's how to interpret them:
X-axis (Number of Sequences Sampled): The total number of sequences sampled (rarefaction depth).
Y-axis (Number of Species/OTUs Observed): The number of distinct species/OTUs observed at each sampling depth.

What to look for?
When the curves plateau, it indicates that most of the species present in the sample have been detected ( the just of the community composition has been found)  and additional sequencing would yield few new species. If many curves plateau, it suggests sufficient sequencing depth for those samples. But you also want to choose a spot that includes most of the samples! 
Rarefy at 40,000?


Making multiple other visualizations to determine where to rarefy.
```{r}
total <- rowSums(seqtab.cleaner)
#total <- rowSums(seqtab.cleaner)
#make some plots to figure out where to rarefy
total.df <- data.frame(total)
ggplot(total.df, aes(x = total)) +
  geom_histogram()

ggplot(total.df, aes(x=1, y=total))+
  geom_boxplot()

ggplot(total.df, aes(x=1, y=total))+
  geom_jitter()+
  scale_y_log10()
```
Other checks to look for where to rarefy?
Histogram: showing distribution of sample reads (should cut out the lowest samples)
Boxplot: Should end up rarefying right below Q1
Jitterplot: least helpful, demonstrates the distribution of sequencing depths across your samples. The y-coordinate represents the total number of reads for each sample, you want them all to be close together and high generally.

```{r}
library(dplyr)
total <- rowSums(seqtab.cleaner)
total.df %>%
  arrange(total) %>%
  ggplot(aes(x=1:nrow(.),y=total))+
  geom_line()  #shows # of samples on x and # of seqs in each sample on y
```
Visualization of sequencing depths across samples. Looking for jumps in your data - probably want to get rid samples with # of reads below large jumps. 
```{r}
total.df %>%
  arrange(total) %>%
  print(20)
```
This code is taking samples and organizing them from lowest to highest # of reads.
Can determine where there are large jumps in your sequencing depths. this is where you make a cutoff. 
There are 2 jumps in my data, so 2 choices:
(1) Cut the first section - anything less than 1500? ** I think we will start with this one 
(2) or could cut off anything less than 11,000?


```{r}
subset(total, total <1500)
```
This code is removing any of the samples with reads less than the sequencing depth you identified above. 

Removed the following:
G5   G6   G7   G9   H1   H2   H3   H4   H5   H6 
1431  587  928 1459 1451  336  290  771  548  419 

```{r}
row.names.remove <- c("G5","G6","G7","G9","H1","H2","H3","H4", "H5", "H6")
seqtab.less <- seqtab.cleaner[!(row.names(seqtab.cleaner) %in% row.names.remove),]
samdf.less <- samdf.cleaner[!(row.names(samdf.cleaner) %in% row.names.remove), ]
dim(samdf.less) 
ps.cleaner <- phyloseq(otu_table(seqtab.less, taxa_are_rows=FALSE), 
                    sample_data(samdf.less),                    
                    tax_table(taxa2))
```
This is removing the samples identified above from the dataframe (now have a new dataframe seqtab.less) and also removing these samples from the metadat (samdf is now samdf.less) 
Then this is summarizing how many samples are left after removing the low sequencing depth ones identified (80) and # of taxonomic levels or columns (6) 


```{r}
seqtab.rare <- rrarefy(seqtab.less,sample=30000) 
```
Now, that the samples with too low of a sequencing depth (below 1500) are cut out, we have rarefy the remaining samples (based on rarefication curve and other figures above) at the best rarefication depth where the majority of samples plateau at (indicating community composition has been reached) - for my figures, this was 30,000

```{r}
rarecurve(seqtab.rare,step=1000,label=FALSE)
#phyloseq object but rarefied
```
Visualizing rarefication curve after 30,000 cut-off was implemented. My reads are now between 1,500-30,000. 

```{r}
ps.rare <- phyloseq(otu_table(seqtab.rare, taxa_are_rows=FALSE), 
                    sample_data(samdf.less), 
                    tax_table(taxa2))
ps.rare
```
16,223 taxa & 80 samples 

```{r}
ps.rare <- prune_taxa(taxa_sums(ps.rare) > 0, ps.rare)
ps.rare
```
Removing missing taxa - taxa that was only present in the samples that we just removed are now also being removed.
Final counts: 15,527 taxa and 80 samples

```{r}
seqtab.rare <- data.frame(otu_table(ps.rare))
samdf.rare <- data.frame(sample_data(ps.rare))
```
Turning updated rarefied OTU table (seqtab.rare) and rarefied metadata (samdf.rare) into dataframes.

Saving files!

saving ps.rare
```{r}
saveRDS(ps.rare, file="MG_16S_ps.rare.RDS")
write.csv(seqtab.rare,file="MG_16S_seqtab.rare.csv")
write.csv(samdf.rare,file="MG_16S_samdf.rare.csv")
```


Saving ps.cleaner
```{r}
ps.cleaner <- readRDS(file="MG_16S_ps.cleaner.RDS")
ps.cleaner
```
16223 taxa and 80 samples 


```{r}
write.csv(seqtab.less, file="MG_16S_seqtab.less.csv")
write.csv(samdf.less, file="MG_16S_samdf.less.csv")
saveRDS(ps.cleaner,file="MG_16S_ps.cleaner.RDS")
saveRDS(seqtab.less,file="MG_16S_seqtab.less.RDS")
saveRDS(samdf.less,file="MG_16S_samdf.less.RDS")
saveRDS(ps.rare, file="MG_16S_ps.rare.RDS")
saveRDS(seqtab.rare, file="MG_16S_seqtab.rare.RDS")
saveRDS(samdf.rare, file="MG_16S_samdf.rare.RDS")
```

Saving more files above!

What are these files and where to use them?? SUMMARY 

* Use ps.cleaner for diversity info: this is otus/asv data pre rarefication but post removal of contaminants and negative controls & low sequencing depths
*seqtab.less: dataframe (as a csv file) of otu/asv data PRE rarefication but post removal of low sequencing depths (1500)
*samdf.less: dataframe (as a csv file) of metadata PRE rarefication but post removal of low sequencing depths (1500)
*seqtab.rare:  a csv file - rarefied data with both lower (1500) and upper (above 30000) sequencing depth ends cut off
*samdf.rare: a csv file - rarefied metadata with both lower (1500) and upper (above 30000) sequencing depth ends cut off
* Use ps.rare for community composition info: this is FINAL rarefied data with both lower (1500) and upper (above 30000) sequencing depth ends cut off - so like final final phyloseq object - but not a dataframe - has both otu/asv info and metadata combined



```{r}
mean(sample_sums(ps.cleaner)) 
sd(sample_sums(ps.cleaner)) 
sum(sample_sums(ps.cleaner)) 
```
[1] 64958.45
[1] 31864.68
[1] 5196676


```{r}
mean(sample_sums(ps.rare)) 
sd(sample_sums(ps.rare))
sum(sample_sums(ps.rare)) 

```

[1] 28456.31
[1] 5328.133
[1] 2276505

checking read depths for all objects :)


PIECRUST PIPELINE - TO LOOK AT FUNCTIONALITY (NOT RNA THOUGH) *I AM NOT GOING TO DO THIS, But this would be helpful if you want more info on the functions of microorganisms withuot looking at RNA 


```{r}
ps.cleaner <- readRDS("MG_16S_ps.cleaner.RDS")
```



